---
title: "Activity Analysis"
author: "Liza Cordero"
date: "March 26, 2016"
output: html_document
---

# Reproducible Research: Course Project 1

## Loading and preprocessing the data

The data in activity.csv is loaded into a data frame by using read.csv

```{r}
# Define the classes of the columns in data for faster reading
cls <- c("integer","character","integer")

# Read the csv file into a data frame
df <- read.csv("activity.csv",head=TRUE, colClasses=cls, na.strings="NA")

head(df)
```

The data is preprocessed for analysis by (1) correcting the type of the date column, (2) making a new data frame that excludes missing values.

```{r}
# Preprocess data: Correct type of date column
df$date <- as.Date(df$date)

# Preprocess data: Exclude the NAs
df_noNA <- subset(df,!is.na(df$steps))
```

## What is mean total number of steps taken per day?

The total number of steps taken per day is calculated by using tapply to get the sum.

```{r}
# Calculate sum of steps taken per day
dailyTotal <- tapply(df_noNA$steps, df_noNA$date, sum, na.rm=TRUE, simplify=T)
head(dailyTotal)
```

A histogram of the total number of steps taken per day is generated.

```{r, echo=TRUE}
# Generate the histogram of the total of daily steps taken
hist(x=dailyTotal,
     col="blue",
     breaks=20,
     xlab="Daily Total Steps Taken",
     ylab="Frequency",
     main="The Distribution of Daily Total Steps Taken")

```

The mean of the daily total is calculated to be 10766.19.

```{r}
mean(dailyTotal)
```

The median of the daily total is calculated to be 10765.

```{r}
median(dailyTotal) 
```

## What is the average daily activity pattern?

A time-series plot is generated to see the average daily activity pattern using tapply to get the mean number of steps per interval.  A data frame is made for the time-series plot.

```{r}
# Calculate the mean number of steps per 5-minute interval
avgInterval <- tapply(df_noNA$steps, df_noNA$interval, mean, na.rm=TRUE, simplify=T)

# Make the data frame for the time-series plot
df_avgInt <- data.frame(interval=as.integer(names(avgInterval)),avg=avgInterval)
```

```{r, echo=TRUE}
# Generate the time series plot to find the average daily activity pattern
with(df_avgInt,
      plot(interval,
            avg,
            type="l",
            xlab="5-Minute Intervals",
            ylab="Average Steps in the Interval - All Days"))
```

The interval with the maximum number of steps is calculated to be the 835th interval with 206 number of steps.

```{r}
# Find the 5-minute interval that contains the maximum number of steps on average per day
max_steps <- max(df_avgInt$avg)
df_avgInt[df_avgInt$avg == max_steps, ]
```

## Imputing missing values

The total number of missing values in the dataset is calculated to be 2304.

```{r}
# Calculate the total number of missing values
sum(is.na(df$steps))
```

The mean for the 5-minute interval is used to substitute for the missing values.  A new data frame is generated with the missing values filled in.

```{r}
# Make a new data frame, d_fill, with the missing values replaced by the mean
df_fill <- df
k <- is.na(df_fill$steps)
avgInterval <- tapply(df_noNA$steps, df_noNA$interval, mean, na.rm=TRUE, simplify=T)
df_fill$steps[k] <- avgInterval[as.character(df_fill$interval[k])]
```

A histogram with the total number of steps taken each day is generated.

```{r, echo=FALSE}
# Calculate the daily sum again with the filled data frame
dailySumFilled <- tapply(df_fill$steps, df_fill$date, sum, na.rm=TRUE, simplify=T)

# Generate the Histogram of daily total steps taken with missing data imputed
hist(x=dailySumFilled,
     col="blue",
     breaks=20,
     xlab="Daily Steps",
     ylab="Frequency",
     main="The Distribution of Daily Total Steps With Missing Data Imputed")

```

The mean based on the new dataset is calculated to be 10766.19

```{r}
# Calculate the mean of the new filled data
mean(dailySumFilled)
```

The median based on the new dataset is calculated to be 10766.19

```{r}
# Calculate the median of the new filleddata
median(dailySumFilled)
```

There is very little difference between the median with the missing values taken out (10765) and the one with the missing values filled in (10766.19). The mean is identical when missing values were taken out (10766.19) and when missing values were filled in (10766.19).  There was no significant change because we used the mean to fill the missing data.

In the histogram, there was a smoothing out of the graph towards the center.

## Are there differences in activity patterns between weekdays and weekends?

A new factor variable was created in the dataset with two levels, weekday and weekend, indicating whether the date is a weekday or a weekend.

```{r}
# Add a column that identifies whether the date is a weekend or weekday to df_fill
is_weekday <- function(d) {
     wd <- weekdays(d)
     ifelse (wd == "Saturday" | wd == "Sunday", "weekend", "weekday")
}

wndx <- sapply(df_fill$date, is_weekday)
df_fill$wk <- as.factor(wndx)
head(df_fill)
```

A panel plot containing 2 time-series plot of the 5-minute interval and the average number of steps averaged across weekends and weekdays is generated from the new data set.

```{r, echo=TRUE}
# Generate a time-series plot of weekend and weekday activity
df_dayOrEnd <- aggregate(steps ~ wk+interval, data=df_fill, FUN=mean)
library(lattice)
xyplot(steps ~ interval | factor(wk),
       layout = c(1, 2),
       xlab="Interval",
       ylab="Number of Steps",
       type="l",
       lty=1,
       data=df_dayOrEnd)
```

Based on the two activity plots, it seems that activities start earlier on weekdays than on weekends. Also the number of steps have a greater peak on weekdays. Weekend activity seems to stay pretty consistent from the 800th to 1750th interval and are greater than weekday activity from th 1250th interval to the 1750th interval.